---
title: "Simulation Study"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(pbapply);library(tidyverse);library(scales);library(forcats)
scaler.maker = function(x = 0, factor = 10) {
  maxima = max(x)
  transform = function(x) -log10(-(x - maxima) + 10^(-factor)) - factor
  inverse = function(x) 10^(-factor) - 10^(-factor - x) + maxima
  breaks = function(x, n = nbreaks) {
    minima = min(x)
    return(inverse(seq(transform(minima), 0, length.out = n)))
  }
  scales::trans_new(
    name = "myscale",
    transform = transform,
    inverse = inverse,
    breaks = breaks,
    domain = c(-Inf, maxima)
  )
}

scale_y_pseudolog = function(factor = 10) {
  scale_y_continuous(
    trans = scaler.maker(factor = factor),
    breaks = c(0, -1e-8, -1e-4, -1e0),
    minor_breaks = 0,
    labels = trans_format(
      function(x) - round(-log10(-x)),
      function(x) ifelse(x == -Inf, "0", math_format(10 ^ .x)(x))
    )
  )
}

head.or.na = function(x, n) {if (is.na(n)) return(NA) else head(x, n)}
tail.1 = function(x) {y = tail(x, 1); if (length(y) == 1) y else NA}
max.or.na = function(x) {x = na.omit(x); if (length(x) == 0) return(NA) else max(x)}
getorNA = function(x, ind) {y = x[ind]; if (length(y) == 1L) y else NA}
```

# Initialize Clusters

```{r}
library(parallel)
library(pbapply)
try(stopCluster(cl))
cl = makeCluster(min(detectCores(), 120))
invisible(clusterEvalQ(cl, {
  rm(list = ls())
  library(excelem)
  # source("Simulation Functions.R", local = TRUE);
  invisible(NULL)
}))
pboptions(use_lb = TRUE, nout = 1)
```

# FA Simulation

```{r}
set.seed(1)
params.FA = expand.grid(
  n = c(1000, 2000, 3000),
  p = c(8, 10),
  q = c(4, 6),
  sd = c(0.5, 1, 2),
  rep = 1:100
)
params.FA$seed = sample.int(2147483647L, nrow(params.FA))

sim.FA = pbapply(cl = cl, simplify = FALSE, params.FA, 1, function(par) {
  dat = gen.data.FA(n = par[['n']], p = par[['p']], q = par[['q']], sd = par[['sd']], seed = par[['seed']])
  dat = scale(dat)
  init = init.par.FA(dat, par[['q']])
  
  aem.FA = tryCatch(accelerated.em(
    dat, init, 
    e.step.FA, m.step.FA, mix.step.FA, loglik.FA,
    iters = 10000, pre.em = 100
  ), error = function(e) e)
  
  em.FA = tryCatch(accelerated.em(
    dat, init, 
    e.step.FA, m.step.FA, mix.step.FA, loglik.FA,
    iters = 10000, pre.em = Inf
  ), error = function(e) e)
  
  aecme.FA = tryCatch(accelerated.em(
    dat, init,
    e.step.FA, cm.step.FA, mix.step.FA, loglik.FA,
    iters = 10000, pre.em = 100
  ), error = function(e) e)

  ecme.FA = tryCatch(accelerated.em(
    dat, init,
    e.step.FA, cm.step.FA, mix.step.FA, loglik.FA,
    iters = 10000, pre.em = Inf
  ), error = function(e) e)
  
  sq.em.FA = tryCatch(squarem.wrapper(
    dat, init,
    e.step.FA, m.step.FA, loglik.FA, fixpt.par.wrap.FA, fixpt.par.unwrap.FA,
    maxiter = 10000, objfn.inc = 0
  ), error = function(e) e)
  
  sq.ecme.FA = tryCatch(squarem.wrapper(
    dat, init,
    e.step.FA, cm.step.FA, loglik.FA, fixpt.par.wrap.FA, fixpt.par.unwrap.FA,
    maxiter = 10000, objfn.inc = 0
  ), error = function(e) e)
  
  sqp.em.FA = tryCatch(squarem.wrapper(
    dat, init,
    e.step.FA, m.step.FA, loglik.FA, fixpt.par.wrap.2.FA, fixpt.par.unwrap.2.FA,
    maxiter = 10000, objfn.inc = 0
  ), error = function(e) e)
  
  sqp.ecme.FA = tryCatch(squarem.wrapper(
    dat, init,
    e.step.FA, cm.step.FA, loglik.FA, fixpt.par.wrap.2.FA, fixpt.par.unwrap.2.FA,
    maxiter = 10000, objfn.inc = 0
  ), error = function(e) e)
  
  list("aEM" = aem.FA, "aECME" = aecme.FA,
       "EM" = em.FA, "ECME" = ecme.FA,
       "sqEM" = sq.em.FA, "sqECME" = sq.ecme.FA,
       "sqpEM" = sqp.em.FA, "sqpECME" = sqp.ecme.FA)
})

# Reduce file size by removing large unused elements
for (i in seq_along(sim.FA)) {
  for (j in seq_along(sim.FA[[i]])) {
    try({sim.FA[[i]][[j]]$update = NULL})
    try({sim.FA[[i]][[j]]$alpha.trace = NULL})
  }
}

save(params.FA, sim.FA, file = format(Sys.time(), "%Y-%m-%d %Hh%Mm%Ss FA.rdata"))
```

# GMM Simulation

```{r}
set.seed(1)
params.GMM = expand.grid(
  n = c(1000, 2000, 3000),
  d = c(2, 3, 4),
  sigma = c(0.5, 1, 2),
  rep = 1:100
)
params.GMM$seed = sample.int(2147483647L, nrow(params.GMM))

sim.GMM = pbapply(cl = cl, simplify = FALSE, params.GMM, 1, function(par) {
  dat = gen.data.GMM(n = par['n'], d = par['d'], sigma = par['sigma'], seed = par['seed'])
  init = init.par.GMM(dat, 2^par['d'])
  
  aem.GMM = tryCatch(accelerated.em(
    dat, init, 
    e.step.GMM, m.step.GMM, mix.step.GMM, loglik.GMM,
    iters = 10000, pre.em = 100
  ), error = function(e) NA)
  
  em.GMM = tryCatch(accelerated.em(
    dat, init, 
    e.step.GMM, m.step.GMM, mix.step.GMM, loglik.GMM,
    iters = 10000, pre.em = Inf
  ), error = function(e) NA)
   
  sq.em.GMM = tryCatch(squarem.wrapper(
    dat, init,
    e.step.GMM, m.step.GMM, loglik.GMM, fixpt.par.wrap.GMM, fixpt.par.unwrap.GMM,
    maxiter = 10000, objfn.inc = 0
  ), error = function(e) e)

  sqp.em.GMM = tryCatch(squarem.wrapper(
    dat, init,
    e.step.GMM, m.step.GMM, loglik.GMM, fixpt.par.wrap.2.GMM, fixpt.par.unwrap.2.GMM,
    maxiter = 10000, objfn.inc = 0
  ), error = function(e) e)
  
  list(EM = em.GMM, aEM = aem.GMM, sqEM = sq.em.GMM, sqpEM = sqp.em.GMM)
})

# Reduce file size by removing large unused elements
for (i in seq_along(sim.GMM)) {
  for (j in seq_along(sim.GMM[[i]])) {
    try({sim.GMM[[i]][[j]]$par$logwts = NULL})
    try({sim.GMM[[i]][[j]]$par$zhat = NULL})
    try({sim.GMM[[i]][[j]]$update = NULL})
    try({sim.GMM[[i]][[j]]$alpha.trace = NULL})
  }
}

save(params.GMM, sim.GMM, file = format(Sys.time(), "%Y-%m-%d %Hh%Mm%Ss GMM.rdata"))
```

# VC Simulation

```{r}
set.seed(1)
params.VC3 = expand.grid(
  groups = c(2, 3, 4),
  nj = 100,
  p = 2,
  q = 2,
  noise.sigmasq = c(0.5, 1, 2),
  rep = 1:100
)
params.VC3$seed = sample.int(2147483647L, nrow(params.VC3))

sim.VC3 = pbapply(cl = cl, simplify = FALSE, params.VC3, 1, function(par) {
  dat = gen.data.VC3(groups = par[['groups']], 
                     nj = rep(par[['nj']], par[['groups']]),
                     p = par[['p']], q = par[['q']],
                     noise.sigmasq = par[['noise.sigmasq']],
                     seed = par[['seed']])
  init = init.par.VC3(dat)
  
  aem.VC3 = tryCatch(accelerated.em(
    dat, init, 
    e.step.VC3, m.step.VC3, mix.step.VC3, loglik.VC3,
    iters = 10000, pre.em = 100
  ), error = function(e) e)
  
  em.VC3 = tryCatch(accelerated.em(
    dat, init, 
    e.step.VC3, m.step.VC3, mix.step.VC3, loglik.VC3,
    iters = 10000, pre.em = Inf
  ), error = function(e) e)
  
  aecme1.VC3 = tryCatch(accelerated.em(
    dat, init,
    e.step.VC3, cm1.step.VC3, mix.step.VC3, loglik.VC3,
    iters = 10000, pre.em = 100
  ), error = function(e) e)
  
  ecme1.VC3 = tryCatch(accelerated.em(
    dat, init,
    e.step.VC3, cm1.step.VC3, mix.step.VC3, loglik.VC3,
    iters = 10000, pre.em = Inf
  ), error = function(e) e)
  
  aecme2.VC3 = tryCatch(accelerated.em(
    dat, init,
    e.step.VC3, cm2.step.VC3, mix.step.VC3, loglik.VC3,
    iters = 10000, pre.em = 100
  ), error = function(e) e)
  
  ecme2.VC3 = tryCatch(accelerated.em(
    dat, init,
    e.step.VC3, cm2.step.VC3, mix.step.VC3, loglik.VC3,
    iters = 10000, pre.em = Inf
  ), error = function(e) e)
    
  sq.em.VC3 = tryCatch(squarem.wrapper(
    dat, init,
    e.step.VC3, m.step.VC3, loglik.VC3, fixpt.par.wrap.VC3, fixpt.par.unwrap.VC3,
    maxiter = 10000, objfn.inc = 0
  ), error = function(e) e)
  
  sq.ecme1.VC3 = tryCatch(squarem.wrapper(
    dat, init,
    e.step.VC3, cm1.step.VC3, loglik.VC3, fixpt.par.wrap.VC3, fixpt.par.unwrap.VC3,
    maxiter = 10000, objfn.inc = 0
  ), error = function(e) e)
  
  sq.ecme2.VC3 = tryCatch(squarem.wrapper(
    dat, init,
    e.step.VC3, cm2.step.VC3, loglik.VC3, fixpt.par.wrap.VC3, fixpt.par.unwrap.VC3,
    maxiter = 10000, objfn.inc = 0
  ), error = function(e) e)
  
  sq2.em.VC3 = tryCatch(squarem.wrapper(
    dat, init,
    e.step.VC3, m.step.VC3, loglik.VC3, fixpt.par.wrap.2.VC3, fixpt.par.unwrap.2.VC3,
    maxiter = 10000, objfn.inc = 0
  ), error = function(e) e)
  
  sq2.ecme1.VC3 = tryCatch(squarem.wrapper(
    dat, init,
    e.step.VC3, cm1.step.VC3, loglik.VC3, fixpt.par.wrap.2.VC3, fixpt.par.unwrap.2.VC3,
    maxiter = 10000, objfn.inc = 0
  ), error = function(e) e)
  
  sq2.ecme2.VC3 = tryCatch(squarem.wrapper(
    dat, init,
    e.step.VC3, cm2.step.VC3, loglik.VC3, fixpt.par.wrap.2.VC3, fixpt.par.unwrap.2.VC3,
    maxiter = 10000, objfn.inc = 0
  ), error = function(e) e)
  
  list("aEM" = aem.VC3, "EM" = em.VC3,
       "aECME1" = aecme1.VC3, "ECME1" = ecme1.VC3,
       "aECME2" = aecme2.VC3, "ECME2" = ecme2.VC3,
       "sqEM" = sq.em.VC3, "sqECME1" = sq.ecme1.VC3, "sqECME2" = sq.ecme2.VC3,
       "sqpEM" = sq2.em.VC3, "sqpECME1" = sq2.ecme1.VC3, "sqpECME2" = sq2.ecme2.VC3)
})

# Reduce file size by removing large unused elements
for (i in seq_along(sim.VC3)) {
  for (j in seq_along(sim.VC3[[i]])) {
    sim.VC3[[i]][[j]]$par[c("XtX", "ZtZ", "XtZ", "Zty", "Xty", "sumXtX", "sumXty", "invsumXtX", "Xbeta", "invD", "shortcut1", "E")] = NULL
    sim.VC3[[i]][[j]][c("update", "alpha.trace")] = NULL
  }
}

save(params.VC3, sim.VC3, file = format(Sys.time(), "%Y-%m-%d %Hh%Mm%Ss VC3.rdata"), compress = 'xz')
```


# GMM Analysis

## Generate data.df

```{r}
data.list = pblapply(1:nrow(params.GMM), function(i) {
  L = sim.GMM[[i]]
  df = tibble(
    id = i,
    ll = lapply(L, function(x) tryCatch(x$loglik.trace, error = function(e) NA)),
    time = lapply(L, function(x) tryCatch(cumsum(x$timing), error = function(e) NA)),
    conv = unname(sapply(L, function(x) tryCatch(isTRUE(x$converged), error = function(e) NA))),
    Method = names(L)
  )
  df %>% bind_cols(params.GMM[i,]) -> df
  return(df)
})
data.df = bind_rows(data.list)
data.df %>% select(-seed) %>% mutate(across(c(n, d), as.integer)) -> data.df

data.df %>%
  filter(!startsWith(Method, "sq")) %>%
  mutate(
    stop.iter = map_int(ll, function(ll) {
      tryCatch({d = diff(ll); x = which(d > 0 & d < 1e-10)[1]; if (is.na(x)) length(ll) else x},
               error = function(e) NA)}),
    conv = stop.iter < map_int(ll, length),
    time = map2(time, stop.iter, head),
    ll = map2(ll, stop.iter, head),
    max.ll = map2_dbl(ll, stop.iter, getorNA),
    stop.time = map2_dbl(time, stop.iter, getorNA)
  ) %>%
  bind_rows(
    data.df %>%
      filter(startsWith(Method, "sq")) %>%
      mutate(
        stop.iter = map_int(ll, length),
        max.ll = map2_dbl(ll, stop.iter, getorNA),
        stop.time = map2_dbl(time, stop.iter, getorNA)
      )
  ) %>%
  group_by(id) %>%
  mutate(id.max.ll = max(max.ll, na.rm = TRUE), 
         max.ll = max.ll - id.max.ll) %>%
  ungroup() %>% 
  mutate(
    ll.mod = map2(ll, id.max.ll, `-`),
    near.optima = max.ll > -1e-8,
    grossly.suboptimal = max.ll < -1e-4
  ) %>% 
  group_by(id) %>% 
  mutate(all.near.optima = all(conv & near.optima)) %>%
  ungroup() %>% 
  mutate(
    Method = case_when(
      startsWith(Method, "a") ~ paste0("Leapfrog-", substr(Method, 2, 100)),
      startsWith(Method, "sqp") ~ paste0("SQUAREM-", substr(Method, 4, 100), "-T"),
      startsWith(Method, "sq") ~ paste0("SQUAREM-", substr(Method, 3, 100)),
      .default = Method
    ),
    type = case_when(
      conv & all.near.optima ~ "1) All Near-Optima",
      conv & near.optima ~ "2) Near-Optima",
      conv & !grossly.suboptimal ~ "3) Sub-Optimal",
      conv & grossly.suboptimal ~ "4) Very Sub-Optimal",
      .default = "5) Did Not Converge"
    )
  ) %>% 
  select(-ll, -id.max.ll, 
         -rep, -conv, -near.optima, 
         -all.near.optima, -grossly.suboptimal) -> data.df.2
```

## Convergence Table and Plot

```{r}
data.df.2 %>% count(type, Method, name = "count") %>% 
  pivot_wider(names_from = Method, values_from = count, values_fill = 0) -> temp

print(temp)
temp %>% select(type, ends_with("EM") | ends_with("EM-T")) %>% knitr::kable(booktabs = TRUE, format = "latex") %>% print
temp %>% select(type, ends_with("ECME") | ends_with("ECME-T")) %>% knitr::kable(booktabs = TRUE, format = "latex") %>% print

data.df.2 %>% 
  mutate(
    Method = fct_reorder(Method, type, function(x) {sum(startsWith(x, "2)") | startsWith(x, "1)"))}, .na_rm = TRUE)
  ) %>% 
  ggplot(aes(x = Method, fill = type)) + 
  geom_bar() + 
  facet_grid(paste0("\u03c3² = ", sigma) ~ paste0("d = ", d)) +
  scale_fill_brewer(palette = "Reds") + 
  labs(x = "", y = "Frequency", fill = "") +
  theme_bw() + 
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  )+
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  coord_flip()

ggsave("t:/gmm_convergence_by_par.pdf", device = cairo_pdf, width = 6, height = 8, dpi = 300)
```

## Wall-Clock Time Boxplot

```{r}
data.df.2 %>% 
  mutate(
    tot.time = map_dbl(time, \(x) {y = tail(x, 1); if (length(y) == 1) y else NA}),
    Method = fct_reorder(factor(Method), -tot.time, mean, .na_rm = TRUE)
  ) %>% 
  group_by(Method) %>% mutate(mean.time = mean(tot.time, na.rm = TRUE)) %>% 
  ggplot(aes(x = Method, y = tot.time, fill = -mean.time)) + 
  geom_violin(scale = "width") + 
  scale_y_log10(labels = math_format(.x)) +
  labs(x = "", y = "Wall-Clock Time (seconds)", fill = "") +
  guides(fill = "none") + theme_bw() + theme() + coord_flip()
ggsave("t:/gmm_time_boxplot.pdf", device = pdf, width = 6, height = 4, dpi = 300)
```

## Wall-Clock Log-Lik Boxplot

```{r}
data.df.2 %>% 
  mutate(
    Method = fct_reorder(factor(Method), max.ll, mean, .na_rm = TRUE)
  ) %>% 
  group_by(Method) %>% mutate(mean.time = mean(max.ll, na.rm = TRUE)) %>% 
  ggplot(aes(x = Method, y = max.ll, fill = mean.time)) + 
  geom_violin(scale = "width") + 
  scale_y_continuous(trans = scaler.maker(0, factor = 10, nbreaks = 5),
                     minor_breaks = 0,
                     breaks = c(0, -1e-8, -1e-4, -1e0),
                     labels = trans_format(function(x) -round(-log10(-x)),
                                           function(x) ifelse(x == -Inf, "0", math_format(10^.x)(x)))) +
  labs(x = "", y = "Log-Likelihood \u0394max", fill = "") +
  guides(fill = "none") + theme_bw() + theme() + coord_flip()
ggsave("t:/gmm_maxll_boxplot.pdf", device = cairo_pdf, width = 6, height = 3, dpi = 300)
```

## Time vs Log-Likelihood Scatterplot

```{r}
data.df.2 %>% 
  mutate(Method = fct_reorder(Method, Method, function(x) {
    y = 1 * grepl("EM(-T|$)", x) + 2 * grepl("ECME1(-T|$)", x) + 3 * grepl("ECME2(-T|$)", x) +
        0.1 * grepl("Leapfrog", x) + 10 * grepl("SQUAREM", x) +
        0.1 * grepl("-T", x)
    mean(y)
  })) %>% 
  ggplot(aes(x = stop.time, y = max.ll, colour = Method)) +
  geom_point(alpha = 0.5) +
  scale_y_continuous(trans = scaler.maker(0, factor = 10, nbreaks = 5),
                   minor_breaks = 0,
                   breaks = c(0, -1e-8, -1e-4, -1e0),
                   labels = trans_format(function(x) -round(-log10(-x)),
                                         function(x) ifelse(x == -Inf, "0", math_format(10^.x)(x)))) +
  scale_x_log10(labels = math_format(.x)) +
  scale_colour_brewer(palette = "Paired") +
  labs(x = "Wall-Clock Time (seconds)", y = "Log-Likelihood Δmax", colour = "") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.margin = margin(0, 0, 0, 0),
    legend.spacing.x = unit(0, "mm"),
    legend.spacing.y = unit(0, "mm")
  ) +
  facet_grid(. ~ case_when(grepl("SQUAREM", Method) ~ "c) SQUAREM",
                         grepl("Leapfrog", Method) ~ "b) Leapfrog",
                         .default = "a) Unaccelerated")) + 
  guides(colour = guide_legend(nrow = 2, override.aes = list(alpha = 1)))
ggsave("t:/gmm_time_vs_deltamax.png", width = 6, height = 4, dpi = 300)
```


# VC3 Analysis

## Generate data.df

```{r}
data.list = pblapply(1:nrow(params.VC3), function(i) {
  L = sim.VC3[[i]]
  df = tibble(
    id = i,
    ll = lapply(L, function(x) tryCatch(x$loglik.trace, error = function(e) NA)),
    time = lapply(L, function(x) tryCatch(cumsum(x$timing), error = function(e) NA)),
    conv = unname(sapply(L, function(x) tryCatch(isTRUE(x$converged), error = function(e) NA))),
    Method = names(L)
  )
  df %>% bind_cols(params.VC3[i,]) -> df
  return(df)
})
data.df = bind_rows(data.list)
data.df %>% select(-seed) %>% mutate(across(c(groups, nj, p, q), as.integer)) -> data.df

data.df %>%
  filter(!startsWith(Method, "sq")) %>%
  mutate(
    stop.iter = map_int(ll, function(ll) {
      tryCatch({d = diff(ll); x = which(d > 0 & d < 1e-10)[1]; if (is.na(x)) length(ll) else x},
               error = function(e) NA)}),
    conv = stop.iter < map_int(ll, length),
    time = map2(time, stop.iter, head),
    ll = map2(ll, stop.iter, head),
    max.ll = map2_dbl(ll, stop.iter, getorNA),
    stop.time = map2_dbl(time, stop.iter, getorNA)
  ) %>%
  bind_rows(
    data.df %>%
      filter(startsWith(Method, "sq")) %>%
      mutate(
        stop.iter = map_int(ll, length),
        max.ll = map2_dbl(ll, stop.iter, getorNA),
        stop.time = map2_dbl(time, stop.iter, getorNA)
      )
  ) %>%
  group_by(id) %>%
  mutate(id.max.ll = max(max.ll, na.rm = TRUE), 
         max.ll = max.ll - id.max.ll) %>%
  ungroup() %>% 
  mutate(
    ll.mod = map2(ll, id.max.ll, `-`),
    near.optima = max.ll > -1e-8,
    grossly.suboptimal = max.ll < -1e-4
  ) %>% 
  group_by(id) %>% 
  mutate(all.near.optima = all(conv & near.optima)) %>%
  ungroup() %>% 
  mutate(
    Method = case_when(
      startsWith(Method, "a") ~ paste0("Leapfrog-", substr(Method, 2, 100)),
      startsWith(Method, "sqp") ~ paste0("SQUAREM-", substr(Method, 4, 100), "-T"),
      startsWith(Method, "sq") ~ paste0("SQUAREM-", substr(Method, 3, 100)),
      .default = Method
    ),
    type = case_when(
      conv & all.near.optima ~ "1) All Near-Optima",
      conv & near.optima ~ "2) Near-Optima",
      conv & !grossly.suboptimal ~ "3) Sub-Optimal",
      conv & grossly.suboptimal ~ "4) Very Sub-Optimal",
      .default = "5) Did Not Converge"
    )
  ) %>% 
  select(-ll, -id.max.ll, 
         -rep, -conv, -near.optima, 
         -all.near.optima, -grossly.suboptimal) -> data.df.2
```

## Convergence Table and Plot

```{r}
data.df.2 %>% count(type, Method, name = "count") %>% 
  pivot_wider(names_from = Method, values_from = count, values_fill = 0) -> temp

print(temp)
temp %>% select(type, ends_with("EM") | ends_with("EM-T")) %>% knitr::kable(booktabs = TRUE, format = "latex") %>% print
temp %>% select(type, ends_with("ECME1") | ends_with("ECME1-T")) %>% knitr::kable(booktabs = TRUE, format = "latex") %>% print
temp %>% select(type, ends_with("ECME2") | ends_with("ECME2-T")) %>% knitr::kable(booktabs = TRUE, format = "latex") %>% print

data.df.2 %>% 
  mutate(
    Method = fct_reorder(Method, type, function(x) {sum(startsWith(x, "2)") | startsWith(x, "1)"))}, .na_rm = TRUE)
  ) %>% 
  ggplot(aes(x = Method, fill = type)) + 
  geom_bar() + 
  facet_grid(paste0("\u03C3² = ", noise.sigmasq) ~ paste0("m = ", groups)) +
  scale_fill_brewer(palette = "Reds") + 
  labs(x = "", y = "Frequency", fill = "") +
  theme_bw() + 
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  )+
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  coord_flip()

ggsave("t:/vc_convergence_by_par.pdf", device = cairo_pdf, width = 6, height = 8, dpi = 300)
```

## Wall-Clock Time Boxplot

```{r}
data.df.2 %>% 
  mutate(
    tot.time = map_dbl(time, \(x) {y = tail(x, 1); if (length(y) == 1) y else NA}),
    Method = fct_reorder(factor(Method), -tot.time, mean, .na_rm = TRUE),
  ) %>% 
  mutate(type = case_when(
    type == "1) All Near-Optima" ~ "Near-Optima",
    type == "2) Near-Optima" ~ "Near-Optima",
    type == "3) Sub-Optimal" ~ "Sub-Optimal",
    type == "4) Very Sub-Optimal" ~ "Sub-Optimal",
    # .default = type
    .default = "Non-Convergent"
  )) %>% 
  group_by(Method) %>% mutate(mean.time = mean(tot.time)) %>% 
  ggplot(aes(x = Method, y = tot.time, fill = -mean.time)) + 
  geom_violin() + 
  scale_y_log10(labels = math_format(.x)) +
  labs(x = "", y = "Wall-Clock Time (seconds)", fill = "") +
  guides(fill = "none") + theme_bw() + theme() + coord_flip()
ggsave("t:/vc_time_boxplot.pdf", device = cairo_pdf, width = 6, height = 3, dpi = 300)
```

## Wall-Clock Log-Lik Boxplot

```{r}
data.df.2 %>% 
  mutate(
    Method = fct_reorder(factor(Method), max.ll, mean, .na_rm = TRUE)
  ) %>% 
  group_by(Method) %>% mutate(mean.time = mean(max.ll)) %>% 
  ggplot(aes(x = Method, y = max.ll, fill = mean.time)) + 
  geom_violin(scale = "width") + 
  scale_y_continuous(trans = scaler.maker(0, factor = 10, nbreaks = 5),
                     minor_breaks = 0,
                     breaks = c(0, -1e-8, -1e-4, -1e0),
                     labels = trans_format(function(x) -round(-log10(-x)),
                                           function(x) ifelse(x == -Inf, "0", math_format(10^.x)(x)))) +
  labs(x = "", y = "Log-Likelihood \u0394max", fill = "") +
  guides(fill = "none") + theme_bw() + theme() + coord_flip()
ggsave("t:/vc_maxll_boxplot.pdf", device = cairo_pdf, width = 6, height = 3, dpi = 300)
```

## Time vs Log-Likelihood Scatterplot

```{r}
data.df.2 %>% 
  mutate(Method = fct_reorder(Method, Method, function(x) {
    y = 1 * grepl("EM(-T|$)", x) + 2 * grepl("ECME1(-T|$)", x) + 3 * grepl("ECME2(-T|$)", x) +
        0.1 * grepl("Leapfrog", x) + 10 * grepl("SQUAREM", x) +
        0.1 * grepl("-T", x)
    mean(y)
  })) %>% 
  ggplot(aes(x = stop.time, y = max.ll, colour = Method)) +
  geom_point(alpha = 0.5) +
  scale_y_continuous(trans = scaler.maker(0, factor = 10, nbreaks = 5),
                   minor_breaks = 0,
                   breaks = c(0, -1e-8, -1e-4, -1e0),
                   labels = trans_format(function(x) -round(-log10(-x)),
                                         function(x) ifelse(x == -Inf, "0", math_format(10^.x)(x)))) +
  scale_x_log10(labels = math_format(.x)) +
  scale_colour_brewer(palette = "Paired") +
  labs(x = "Wall-Clock Time (seconds)", y = "Log-Likelihood Δmax", colour = "") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.margin = margin(0, 0, 0, 0),
    legend.spacing.x = unit(0, "mm"),
    legend.spacing.y = unit(0, "mm")
  ) +
  facet_grid(. ~ case_when(grepl("SQUAREM", Method) ~ "c) SQUAREM",
                         grepl("Leapfrog", Method) ~ "b) Leapfrog",
                         .default = "a) Unaccelerated")) + 
  guides(colour = guide_legend(nrow = 4, override.aes = list(alpha = 1)))
ggsave("t:/vc_time_vs_deltamax.png", width = 6, height = 4, dpi = 300)
```


# FA Analysis

## Generate data.df

```{r}
is.heywood = function(par) {
  tryCatch({min(diag(par$BBt), par$D)}, error = function(e) NA)
}

data.list = pblapply(1:nrow(params.FA), function(i) {
  L = sim.FA[[i]]
  df = tibble(
    id = i,
    ll = lapply(L, function(x) tryCatch(x$loglik.trace, error = function(e) NA)),
    time = lapply(L, function(x) tryCatch(cumsum(x$timing), error = function(e) NA)),
    conv = unname(sapply(L, function(x) tryCatch(isTRUE(x$converged), error = function(e) NA))),
    heywood = sapply(L, function(x) is.heywood(x$par)),
    Method = names(L)
  )
  df %>% bind_cols(params.FA[i,]) -> df
  return(df)
})
data.df = bind_rows(data.list)
data.df %>% select(-seed) %>% mutate(across(c(n, p, q), as.integer)) -> data.df

data.df %>%
  filter(!startsWith(Method, "sq")) %>%
  mutate(
    stop.iter = map_int(ll, function(ll) {
      tryCatch({d = diff(ll); x = which(d > 0 & d < 1e-10)[1]; if (is.na(x)) length(ll) else x},
               error = function(e) NA)}),
    conv = stop.iter < map_int(ll, length),
    time = map2(time, stop.iter, head),
    ll = map2(ll, stop.iter, head),
    max.ll = map2_dbl(ll, stop.iter, getorNA),
    stop.time = map2_dbl(time, stop.iter, getorNA)
  ) %>%
  bind_rows(
    data.df %>%
      filter(startsWith(Method, "sq")) %>%
      mutate(
        stop.iter = map_int(ll, length),
        max.ll = map2_dbl(ll, stop.iter, getorNA),
        stop.time = map2_dbl(time, stop.iter, getorNA)
      )
  ) %>%
  group_by(id) %>%
  mutate(id.max.ll = max(max.ll, na.rm = TRUE), 
         max.ll = max.ll - id.max.ll) %>%
  ungroup() %>% 
  mutate(
    ll.mod = map2(ll, id.max.ll, `-`),
    near.optima = max.ll > -1e-8,
    grossly.suboptimal = max.ll < -1e-4
  ) %>% 
  group_by(id) %>% 
  mutate(all.near.optima = all(conv & near.optima)) %>%
  ungroup() %>% 
  mutate(
    Method = case_when(
      startsWith(Method, "a") ~ paste0("Leapfrog-", substr(Method, 2, 100)),
      startsWith(Method, "sqp") ~ paste0("SQUAREM-", substr(Method, 4, 100), "-T"),
      startsWith(Method, "sq") ~ paste0("SQUAREM-", substr(Method, 3, 100)),
      .default = Method
    ),
    type = case_when(
      conv & all.near.optima ~ "1) All Near-Optima",
      conv & near.optima ~ "2) Near-Optima",
      conv & !grossly.suboptimal ~ "3) Sub-Optimal",
      conv & grossly.suboptimal ~ "4) Very Sub-Optimal",
      .default = "5) Did Not Converge"
    )
  ) %>% 
  select(-ll, -id.max.ll, 
         -rep, -conv, -near.optima, 
         -all.near.optima, -grossly.suboptimal) -> data.df.2
```

## Convergence Table and Plot

```{r}
data.df.2 %>% count(type, Method, name = "count") %>% 
  pivot_wider(names_from = Method, values_from = count, values_fill = 0) -> temp

print(temp)
temp %>% select(type, ends_with("EM") | ends_with("EM-T")) %>% knitr::kable(booktabs = TRUE, format = "latex") %>% print
temp %>% select(type, ends_with("ECME") | ends_with("ECME-T")) %>% knitr::kable(booktabs = TRUE, format = "latex") %>% print

data.df.2 %>% 
  mutate(
    Method = fct_reorder(Method, type, function(x) {sum(startsWith(x, "2)") | startsWith(x, "1)"))}, .na_rm = TRUE)
  ) %>% 
  ggplot(aes(x = Method, fill = type)) + 
  geom_bar() + 
  facet_grid(paste0("p = ", p, ", q = ", q) ~ paste0("\u03c3² = ", sd)) +
  scale_fill_brewer(palette = "Reds") + 
  labs(x = "", y = "Frequency", fill = "") +
  theme_bw() + 
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  )+
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  coord_flip()

ggsave("t:/fa_convergence_by_par.pdf", device = cairo_pdf, width = 6, height = 8, dpi = 300)
```

## Wall-Clock Time Boxplot

```{r}
data.df.2 %>% 
  mutate(
    tot.time = map_dbl(time, \(x) {y = tail(x, 1); if (length(y) == 1) y else NA}),
    Method = fct_reorder(factor(Method), -tot.time, mean, .na_rm = TRUE)
  ) %>% 
  group_by(Method) %>% mutate(mean.time = mean(tot.time)) %>% 
  ggplot(aes(x = Method, y = tot.time, fill = -mean.time)) + 
  geom_violin(scale = "width") + 
  scale_y_log10(labels = math_format(.x)) +
  labs(x = "", y = "Wall-Clock Time (seconds)", fill = "") +
  guides(fill = "none") + theme_bw() + theme() + coord_flip()
ggsave("t:/fa_time_boxplot.pdf", device = cairo_pdf, width = 6, height = 3, dpi = 300)
```

## Wall-Clock Log-Lik Boxplot

```{r}
data.df.2 %>% 
  mutate(
    Method = fct_reorder(factor(Method), max.ll, mean, .na_rm = TRUE)
  ) %>% 
  group_by(Method) %>% mutate(mean.time = mean(max.ll)) %>% 
  ggplot(aes(x = Method, y = max.ll, fill = mean.time)) + 
  geom_violin(scale = "width") + 
  scale_y_continuous(trans = scaler.maker(0, factor = 10, nbreaks = 5),
                     minor_breaks = 0,
                     breaks = c(0, -1e-8, -1e-4, -1e0),
                     labels = trans_format(function(x) -round(-log10(-x)),
                                           function(x) ifelse(x == -Inf, "0", math_format(10^.x)(x)))) +
  labs(x = "", y = "Log-Likelihood \u0394max", fill = "") +
  guides(fill = "none") + theme_bw() + theme() + coord_flip()
ggsave("t:/fa_maxll_boxplot.pdf", device = cairo_pdf, width = 6, height = 3, dpi = 300)
```

## Time vs Log-Likelihood Scatterplot

```{r}
data.df.2 %>% 
  mutate(Method = fct_reorder(Method, Method, function(x) {
    y = 1 * grepl("EM(-T|$)", x) + 2 * grepl("ECME1(-T|$)", x) + 3 * grepl("ECME2(-T|$)", x) +
        0.1 * grepl("Leapfrog", x) + 10 * grepl("SQUAREM", x) +
        0.1 * grepl("-T", x)
    mean(y)
  })) %>% 
  ggplot(aes(x = stop.time, y = max.ll, colour = Method)) +
  geom_point(alpha = 0.5) +
  scale_y_continuous(trans = scaler.maker(0, factor = 10, nbreaks = 5),
                   minor_breaks = 0,
                   breaks = c(0, -1e-8, -1e-4, -1e0),
                   labels = trans_format(function(x) -round(-log10(-x)),
                                         function(x) ifelse(x == -Inf, "0", math_format(10^.x)(x)))) +
  scale_x_log10(labels = math_format(.x)) +
  scale_colour_brewer(palette = "Paired") +
  labs(x = "Wall-Clock Time (seconds)", y = "Log-Likelihood Δmax", colour = "") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.margin = margin(0, 0, 0, 0),
    legend.spacing.x = unit(0, "mm"),
    legend.spacing.y = unit(0, "mm")
  ) +
  facet_grid(. ~ case_when(grepl("SQUAREM", Method) ~ "c) SQUAREM",
                         grepl("Leapfrog", Method) ~ "b) Leapfrog",
                         .default = "a) Unaccelerated")) + 
  guides(colour = guide_legend(nrow = 2, override.aes = list(alpha = 1)))
ggsave("t:/fa_time_vs_deltamax.png", width = 6, height = 4, dpi = 300)
```